---
title: "Example of using mitch package for Infinium methylation analysis - Schizophrenia EWAS"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
bibliography: references.bib
csl: epigenetics.csl
---

## Introduction

Here we are looking at putting together the easiest possible workflow for the package.

Source code: https://github.com/markziemann/gmea/blob/main/example_workflow/example_workflow.Rmd

## Requirements

Load packages.

Important: ensure that the mitch version used is 1.15.1 or higher.

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("HGNChelper")
  library("mitch")
  library("kableExtra")
})

if( packageVersion("mitch") < "1.15.1") {
  warning("This workflow requires mitch version 1.15.1 or higher")
}

```

## Load pathways

Gene ontologies were downloaded in GMT format from MSigDB on 15th Jan 2024[@Liberzon2015-um;@The_Gene_Ontology_Consortium2023-fm].
The GMT file is read into R using the mitch function `gmt_import()`.

```{r,loadpathways}

gene_sets <- gmt_import("c5.go.v2023.2.Hs.symbols.gmt")

```

## Curate the annotation

One of the critical parts of this workflow is the establishment of probe-gene relationships.
This controls how the probe data is aggregated to make the gene level scores.

As these annotations are several years old, many of the annotated gene names are no longer
current.
To remedy this, the gene names are screened with the HGNChelper package and any defunct
symbols get updated to the newer gene name, so they will be recognised properly in the
gene sets.

```{r,anno1}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt1 <- stack(gp2)
colnames(gt1) <- c("gene","probe")
gt1$probe <- as.character(gt1$probe)
dim(gt1)
str(gt1)
length(unique(gt1$gene))


if (! file.exists("new.hgnc.table.rds")) {
  new.hgnc.table <- getCurrentHumanMap()
  saveRDS(new.hgnc.table, "new.hgnc.table.rds")
}

new.hgnc.table <- readRDS("new.hgnc.table.rds")
fix <- checkGeneSymbols(gt1$gene,map=new.hgnc.table)
fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
length(unique(fix2$x))
gt1$gene <- fix$Suggested.Symbol

head(gt1)

```

## Mitch pipeline

The first part is to import the data into mitch.

This only works properly when mitch's mapGeneIds function uses `mean` to aggregate (line 69 of mitch.R).

Data files:

* EWAS_ageofonset_asrb_epic.txt.gz

* EWAS_clozapine_asrb_epic.txt.gz

* EWAS_cognitivestatus_asrb_epic.txt.gz

* EWAS_gafscore_asrb_epic.txt.gz

* EWAS_prs_asrb_epic.txt.gz

```{r,load}

ageonset <- read.table("ASRB_EWAS/EWAS_ageofonset_asrb_epic.fmt.txt",sep="\t",header=TRUE)
colnames(ageonset) <- c("probe","ageonset")
rownames(ageonset) <- ageonset$probe ; ageonset$probe=NULL
iageonset <- -ageonset

cloz <- read.table("ASRB_EWAS/EWAS_clozapine_asrb_epic.fmt.txt",sep="\t",header=TRUE)
colnames(cloz) <- c("probe","cloz")
rownames(cloz) <- cloz$probe ; cloz$probe=NULL

cognit <- read.table("ASRB_EWAS/EWAS_cognitivestatus_asrb_epic.fmt.txt",sep="\t",header=TRUE)
colnames(cognit) <- c("probe","cognit")
rownames(cognit) <- cognit$probe ; cognit$probe=NULL

gaf <- read.table("ASRB_EWAS/EWAS_gafscore_asrb_epic.fmt.txt",sep="\t",header=TRUE)
colnames(gaf) <- c("probe","gaf")
rownames(gaf) <- gaf$probe ; gaf$probe=NULL
igaf <- -gaf

prs <- read.table("ASRB_EWAS/EWAS_prs_asrb_epic.fmt.txt",sep="\t",header=TRUE)
colnames(prs) <- c("probe","prs")
rownames(prs) <- prs$probe ; prs$probe=NULL

```

Import data.

```{r,mergit2}

mylist2 <- list("iageonset"=iageonset,"cloz"=cloz,"cognit"=cognit,"igaf"=igaf,"prs"=prs)

mm <- mitch_import(x=mylist2, DEtype="prescored",geneTable=gt1)

```

Multimitch

```{r,multimitch}

mmres <- mitch_calc(x=mm,genesets=gene_sets,minsetsize=5,cores=16,resrows=50,priority="effect")

if (! file.exists("multires.html")){
  mitch_report(res=mmres,outfile="multires.html")
}

mmrestop <- mmres$enrichment_result
mmrestop <-  mmrestop[order(mmrestop$pMANOVA),]

head(mmrestop,30) %>%
  kbl(caption = "Top results in multi by p-value") %>%
  kable_paper("hover", full_width = F)

mmrestop <- subset(mmrestop,p.adjustMANOVA<0.05)

mmrestop <- mmrestop[order(-abs(mmrestop$s.dist)),]

head(mmrestop,30) %>%
  kbl(caption = "Top results in multi by effect size") %>%
  kable_paper("hover", full_width = F)


```

Single mitch for iageonset, cloz, cognit, igaf and prs.

```{r,mitchsingle}

# age of onset
iageonset_g <- mitch_import(x=iageonset,DEtype="prescored",geneTable=gt1)
m_iageonset <- mitch_calc(iageonset_g,genesets=gene_sets,cores=16, minsetsize=5, resrows=50, priority="effect")
head(m_iageonset$enrichment_result,10) %>% kbl(caption = "Top 10 effect size") %>% kable_paper("hover", full_width = F)
top <- m_iageonset$enrichment_result
top <- top[order(top$pANOVA),]
head(top,10) %>% kbl(caption = "Top 10 significance") %>% kable_paper("hover", full_width = F)
nrow(subset(top,p.adjustANOVA<0.05))

# cloz
cloz_g <- mitch_import(x=cloz,DEtype="prescored",geneTable=gt1)
m_cloz <- mitch_calc(cloz_g,genesets=gene_sets,cores=16, minsetsize=5, resrows=50, priority="effect")
head(m_cloz$enrichment_result,10) %>% kbl(caption = "Top 10 effect size") %>% kable_paper("hover", full_width = F)
top <- m_cloz$enrichment_result
top <- top[order(top$pANOVA),]
head(top,10) %>% kbl(caption = "Top 10 significance") %>% kable_paper("hover", full_width = F)
nrow(subset(top,p.adjustANOVA<0.05))

# cognit
cognit_g <- mitch_import(x=cognit,DEtype="prescored",geneTable=gt1)
m_cognit <- mitch_calc(cognit_g,genesets=gene_sets,cores=16, minsetsize=5, resrows=50, priority="effect")
head(m_cognit$enrichment_result,10) %>% kbl(caption = "Top 10 effect size") %>% kable_paper("hover", full_width = F)
top <- m_cognit$enrichment_result
top <- top[order(top$pANOVA),]
head(top,10) %>% kbl(caption = "Top 10 significance") %>% kable_paper("hover", full_width = F)
nrow(subset(top,p.adjustANOVA<0.05))

# igaf
igaf_g <- mitch_import(x=igaf,DEtype="prescored",geneTable=gt1)
m_igaf <- mitch_calc(igaf_g,genesets=gene_sets,cores=16, minsetsize=5, resrows=50, priority="effect")
head(m_igaf$enrichment_result,10) %>% kbl(caption = "Top 10 effect size") %>% kable_paper("hover", full_width = F)
top <- m_igaf$enrichment_result
top <- top[order(top$pANOVA),]
head(top) %>% kbl(caption = "Top 10 significance") %>% kable_paper("hover", full_width = F)
nrow(subset(top,p.adjustANOVA<0.05))

# prs
prs_g <- mitch_import(x=prs,DEtype="prescored",geneTable=gt1)
m_prs <- mitch_calc(prs_g,genesets=gene_sets,cores=16, minsetsize=5, resrows=50, priority="effect")
head(m_prs$enrichment_result,10) %>% kbl(caption = "Top 10 effect size") %>% kable_paper("hover", full_width = F)
top <- m_prs$enrichment_result
top <- top[order(top$pANOVA),]
head(top,10) %>% kbl(caption = "Top 10 significance") %>% kable_paper("hover", full_width = F)
nrow(subset(top,p.adjustANOVA<0.05))

```

Now make a barplot of these top findings.

```{r, barplot}

par(mar=c(5,27,5,3))

# iageonset
top <- m_iageonset$enrichment_result
top <- top[order(top$pANOVA),]
top <- subset(top,p.adjustANOVA<0.05)
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.7,main="Age of onset",xlab="ES")
grid()

# cloz
top <- m_cloz$enrichment_result
top <- top[order(top$pANOVA),]
top <- subset(top,p.adjustANOVA<0.05)
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.7,main="Clozapine",xlab="ES")
grid()

# cognitive
top <- m_cognit$enrichment_result
top <- top[order(top$pANOVA),]
top <- subset(top,p.adjustANOVA<0.05)
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.6,main="Cognitive function",xlab="ES")
grid()

# igaf
top <- m_igaf$enrichment_result
top <- top[order(top$pANOVA),]
top <- subset(top,p.adjustANOVA<0.05)
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.6,main="iGAF score",xlab="ES")
grid()

# prs
top <- m_prs$enrichment_result
top <- top[order(top$pANOVA),]
top <- subset(top,p.adjustANOVA<0.05)
up <- head(subset(top,s.dist>0),20)
dn <- head(subset(top,s.dist<0),20)
top <- rbind(up,dn)
vec=top$s.dist
names(vec)=top$set
names(vec) <- gsub("_"," ",names(vec))
vec <- vec[order(vec)]
barplot(abs(vec),col=sign(-vec)+3,horiz=TRUE,las=1,cex.names=0.7,main="PRS",xlab="ES")
grid()

par( mar = c(5.1, 4.1, 4.1, 2.1) )

```

## Reports

```{r,mitchreports}

if(!file.exists("mitch_iageonset.html")) {
  mitch_report(res=m_iageonset,outfile="mitch_iageonset.html")
}

if(!file.exists("mitch_cloz.html")) {
  mitch_report(res=m_cloz,outfile="mitch_cloz.html")
}

if(!file.exists("mitch_cognit.html")) {
  mitch_report(res=m_cognit,outfile="mitch_cognit.html")
}

if(!file.exists("mitch_igaf.html")) {
  mitch_report(res=m_igaf,outfile="mitch_igaf.html")
}

if(!file.exists("mitch_prs.html")) {
  mitch_report(res=m_prs,outfile="mitch_prs.html")
}

```
## Session information

```{r,save}

sessionInfo()

```

